---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

temp_dir = tempdir(check = TRUE)

pkg_has_vignettes_and_test = function(pkg_name, pkg_dir = temp_dir) {
  pkg_archive = download.packages(pkg_name, destdir = pkg_dir,
                                  repos = "https://cran.r-project.org")
  
  files_list = untar(pkg_archive[1, 2], list = TRUE)
  
  has_vignette = paste0(pkg_archive[1, 1], "/vignettes/") %in% files_list

  has_tests = paste0(pkg_archive[1, 1], "/tests/") %in% files_list

  c(ifelse(has_vignette, "✅", "❌"), ifelse(has_tests, "✅", "❌"))
}
```

# fundiversity

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html#maturing)
[![R build status](https://github.com/Bisaloo/fundiversity/workflows/R-CMD-check/badge.svg)](https://github.com/Bisaloo/fundiversity/actions)
[![Coverage Status](https://codecov.io/gh/Bisaloo/fundiversity/branch/master/graph/badge.svg?token=HR4YH118VT)](https://codecov.io/gh/Bisaloo/fundiversity)
<!-- badges: end -->

`fundiversity` provides a lightweight package to compute common functional
diversity indices. To a get a glimpse of what `fundiversity` can do refer to the
[introductory vignette](https://bisaloo.github.io/fundiversity/articles/fundiversity.html).
The package is built using clear, public 
[design principles](https://bisaloo.github.io/fundiversity/articles/design-principles.html)
inspired from our own experience and user feedback.

## Installation

You can install the stable version from CRAN with:

```{r, eval = FALSE}
install.packages("fundiversity")
```

Alternatively, you can install the development version with:

```{r eval = FALSE}
install.packages("fundiversity", repos = "https://bisaloo.r-universe.dev")
```

## Examples

`fundiversity` lets you compute six functional diversity indices: Functional Richness with `fd_fric()`, intersection with between convex hulls with `fd_fric_intersect()`, Functional Divergence with
`fd_fdiv()`, Rao's Quadratic Entropy with `fd_raoq()`, Functional Dispersion with `fd_fdis()` and Functional Evenness with `fd_feve()`. You can have a brief overview of the indices in the [introductory vignette](https://bisaloo.github.io/fundiversity/articles/fundiversity.html).

All indices can be computed either using global trait data or at the site-level:

```{r example}
library(fundiversity)

# Get trait data included in the package
data("traits_birds")

# Compute Functional Richness of all birds included
fd_fric(traits_birds)

# Compute Functional Divergence
fd_fdiv(traits_birds)

# Compute Rao's Quadratic Entropy
fd_raoq(traits_birds)

# Compute Functional Dispersion
fd_fdis(traits_birds)

# Compute Functional Evenness
fd_feve(traits_birds)
```

To compute Rao's Quadratic Entropy, the user can also provide a distance matrix between species directly:

```{r rao-distance-matrix}
dist_traits_birds = as.matrix(dist(traits_birds))

fd_raoq(traits = NULL, dist_matrix = dist_traits_birds)
```

## Function Summary

```{r child="man/rmdchunks/_fundiversity_functions.Rmd"}
```

## Parallelization

Thanks to the `future.apply` package, all functions (except `fd_raoq()`) within `fundiversity` support parallelization through the `future` backend. To toggle parallelization follow the [`future` syntax](https://cran.r-project.org/package=future):

```{r future-syntax, render=FALSE}
future::plan(future::multisession)
fd_fdiv(traits_birds)
```

For more details please refer to the [parallelization vignette](https://bisaloo.github.io/fundiversity/articles/parallel.html) or use `vignette("parallel", package = "fundiversity")` within R.

## Available functional diversity indices

According to Pavoine & Bonsall (2011) classification, functional diversity indices can be classified in three "domains" that assess different properties of the functional space: richness, divergence, and regularity.
`fundiversity` provides function to compute indices that assess this three facets at the site scale:

|               Scale                |                                      Richness                                            |                   Divergence                                                                                                                                                           |        Evenness                                                                        |
|------------------------------------|------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|----------------------------------------------------------------------------------------|
| α-diversity<br />(= among sites)   | FRic with [`fd_fric()`](https://bisaloo.github.io/fundiversity/reference/fd_fric.html)   | FDiv with [`fd_fdiv()`](https://bisaloo.github.io/fundiversity/reference/fd_fdiv.html)<br />Rao's QE with [`fd_raoq()`](https://bisaloo.github.io/fundiversity/reference/fd_raoq.html)<br />FDis with [`fd_fdis()`](https://bisaloo.github.io/fundiversity/reference/fd_fdis.html) | FEve with [`fd_feve()`](https://bisaloo.github.io/fundiversity/reference/fd_feve.html) |
| β-diversity<br />(= between sites) | FRic pairwise intersection with [`fd_fric_intersect()`](https://bisaloo.github.io/fundiversity/reference/fd_fric_intersect.html)<br />alternatives available in `betapart` | available in `entropart`, `betapart` or `hillR`                                                               | available in `BAT`    |
 
## Related Packages

```{r pkg-vign-tests, include = FALSE}
fd          = pkg_has_vignettes_and_test("FD")
adiv        = pkg_has_vignettes_and_test("adiv")
betapart    = pkg_has_vignettes_and_test("betapart")
entropart   = pkg_has_vignettes_and_test("entropart")
hillr       = pkg_has_vignettes_and_test("hillR")
vegan       = pkg_has_vignettes_and_test("vegan")
tpd         = pkg_has_vignettes_and_test("TPD")
hypervolume = pkg_has_vignettes_and_test("hypervolume")
bat         = pkg_has_vignettes_and_test("BAT")
```

Several other packages exist that compute functional diversity indices. We have performed a [performance comparison](https://bisaloo.github.io/fundiversity/articles/performance.html) between packages. We here mention some of them (but do not mention the wrappers around these packages): 

| Package Name                                          |                     Indices included                        |    Has vignettes   |      Has tests     | On GitHub | On CRAN (last updated)                                    |
|-------------------------------------------------------|-------------------------------------------------------------|--------------------|--------------------|-----------|-----------------------------------------------------------|
| [`FD`](https://github.com/cran/FD)                    | FRic, FDiv, FDis, FEve, Rao's QE, Functional Group Richness |`r fd[[1]]`         |`r fd[[2]]`         | :x:       | ![](https://www.r-pkg.org/badges/last-release/FD)         |
| [`adiv`](https://github.com/cran/adiv)                | Functional Entropy, Functional Redundancy                   |`r adiv[[1]]`       |`r adiv[[2]]`       | :x:       | ![](https://www.r-pkg.org/badges/last-release/adiv)       |
| [`betapart`](https://github.com/cran/betapart)        | Functional β-diversity                                      |`r betapart[[1]]`   |`r betapart[[2]]`   | :x:       | ![](https://www.r-pkg.org/badges/last-release/betapart)   |
| [`entropart`](https://github.com/EricMarcon/entropart)| Functional Entropy                                          |`r entropart[[1]]`  |`r entropart[[2]]`  | ✅        | ![](https://www.r-pkg.org/badges/last-release/entropart)  |  
| [`hillR`](https://github.com/daijiang/hillR)          | Functional Diversity Hill Number                            |`r hillr[[1]]`      |`r hillr[[2]]`      | ✅        | ![](https://www.r-pkg.org/badges/last-release/hillR)      |    
| [`vegan`](https://github.com/vegandevs/vegan)         | Only dendrogram-based FD (`treedive()`)                     |`r vegan[[1]]`      |`r vegan[[2]]`      | ✅        | ![](https://www.r-pkg.org/badges/last-release/vegan)      |                 
| [`TPD`](https://github.com/cran/TPD)                  | FRic, FDiv, FEve but for probability distributions          |`r tpd[[1]]`        |`r tpd[[2]]`        | :x:       | ![](https://www.r-pkg.org/badges/last-release/TPD)        |
| [`hypervolume`](https://github.com/cran/hypervolume)  | Hypervolume measure functional diversity (~FRic)            |`r hypervolume[[1]]`|`r hypervolume[[2]]`| ✅        | ![](https://www.r-pkg.org/badges/last-release/hypervolume)|
| [`BAT`](https://github.com/cardosopmb/BAT)            | β-, Richness, divergence, and evenness with hypervolumes    |`r bat[[1]]`        |`r bat[[2]]`        | ✅        | ![](https://www.r-pkg.org/badges/last-release/BAT)        |  

---
