---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fundiversity

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
<!-- badges: end -->

The goal of fundiversity is to provide a package to compute common functional
diversity indices. The package is built using clear, public 
[design principles](https://github.com/Bisaloo/fundiversity/wiki/Design-principles)
inspired from our own experience and user feedback. We also strive to use the
latest good practice in software and R package development.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

```{r eval = FALSE}
# install.packages("devtools")
devtools::install_github("Bisaloo/fundiversity")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(fundiversity)
## basic example code
```
