---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fundiversity

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
<!-- badges: end -->

The goal of fundiversity is to provide a package to compute common functional
diversity indices. The package is built using clear, public 
[design principles](https://github.com/Bisaloo/fundiversity/wiki/Design-principles)
inspired from our own experience and user feedback. We also strive to use the
latest good practice in software and R package development.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

```{r eval = FALSE}
# install.packages("devtools")
devtools::install_github("Bisaloo/fundiversity")
```

## Example

You can compute functional diversity indices through the functions `fd_fric()`,
`fd_fdiv()`, and `fd_rao()`:

```{r example}
library(fundiversity)

# Get trait data included in the package
data("traits_birds")

# Compute Functional Richness of all birds included
fd_fric(traits_birds)
```


## Related Packages

| Package Name                                          |                    Indices included                       |     Has tests    | On GitHub        | On CRAN | Last updated (CRAN) |
|-------------------------------------------------------|-----------------------------------------------------------|------------------|------------------|---------|---------------------|
| [`FD`](https://github.com/cran/FD)                    |FRic, FDiv, FDis, FEve, Rao's QE, Functional Group Richness|:x:               |:x:               |:white_check_mark:         |2014-08-19           |
| [`adiv`](https://github.com/cran/adiv)                |Functional Entropy, Functional Redundancy                       |:x:               |:x:               |:white_check_mark:         |2020-08-26           |
| [`betapart`](https://github.com/cran/betapart)        |Functional β-diversity                                     |:x:               |:x:               |:white_check_mark:         |2020-09-09           |
| [`entropart`](https://github.com/EricMarcon/entropart)|Functional Entropy                                         |:white_check_mark:|:white_check_mark:|:white_check_mark:        |2020-01-22           |
| [`hillR`](https://github.com/daijiang/hillR)          |Functional Diversity Hill Number                           |:white_check_mark:|:white_check_mark:|:white_check_mark:        |2020-07-07           |
| [`fundiv`](https://github.com/ibartomeus/fundiv)      |FRic, FDiv, FDis, FEve, Rao's QE, dendrogram FD            |:x:               |:white_check_mark:|:x:                                          |-                    | 
| [`vegan`](https://github.com/vegandevs/vegan)         |Only dendrogram-based FD (`treedive()`)                                  |:white_check_mark:|:white_check_mark:|:white_check_mark:        |2019-09-01           |              
| [`TPD`](https://github.com/cran/TPD)                  |FRic, FDiv, FEve but for probability distributions         |:x:              |:x:                |:white_check_mark:         |2019-07-02           |
