---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

temp_dir = tempdir(check = TRUE)

pkg_has_tests = function(pkg_name, pkg_dir = temp_dir) {
  pkg_archive = download.packages(pkg_name, destdir = pkg_dir,
                                  repos = "https://cran.r-project.org")
  
  files_list = untar(pkg_archive[1, 2], list = TRUE)
  
  has_tests = paste0(pkg_archive[1, 1], "/tests/") %in% files_list
  
  ifelse(has_tests, ":white_check_mark:", ":x:")
}
```

# fundiversity

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
<!-- badges: end -->

The goal of fundiversity is to provide a package to compute common functional
diversity indices. The package is built using clear, public 
[design principles](https://github.com/Bisaloo/fundiversity/wiki/Design-principles)
inspired from our own experience and user feedback. We also strive to use the
latest good practice in software and R package development.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

```{r eval = FALSE}
# install.packages("devtools")
devtools::install_github("Bisaloo/fundiversity")
```

## Example

You can compute functional diversity indices through the functions `fd_fric()`,
`fd_fdiv()`, and `fd_rao()`:

```{r example}
library(fundiversity)

# Get trait data included in the package
data("traits_birds")

# Compute Functional Richness of all birds included
fd_fric(traits_birds)
```


## Related Packages

We only mention here packages that propose original indices and not wrappers around these packages.

| Package Name                                          |                     Indices included                        |           Has tests          |     On GitHub      | On CRAN (last updated) |
|-------------------------------------------------------|-------------------------------------------------------------|------------------------------|--------------------|------------------------------|
| [`FD`](https://github.com/cran/FD)                    | FRic, FDiv, FDis, FEve, Rao's QE, Functional Group Richness |`r pkg_has_tests("FD")`       | :x:                | ![](https://www.r-pkg.org/badges/last-release/FD)  |
| [`adiv`](https://github.com/cran/adiv)                | Functional Entropy, Functional Redundancy                   |`r pkg_has_tests("adiv")`     | :x:                | ![](https://www.r-pkg.org/badges/last-release/adiv)|
| [`betapart`](https://github.com/cran/betapart)        | Functional β-diversity                                      |`r pkg_has_tests("betapart")` | :x:                | ![](https://www.r-pkg.org/badges/last-release/betapart)|
| [`entropart`](https://github.com/EricMarcon/entropart)| Functional Entropy                                          |`r pkg_has_tests("entropart")`| :white_check_mark: | ![](https://www.r-pkg.org/badges/last-release/entropart)|
| [`hillR`](https://github.com/daijiang/hillR)          | Functional Diversity Hill Number                            |`r pkg_has_tests("hillR")`    | :white_check_mark: | ![](https://www.r-pkg.org/badges/last-release/hillR)|
| [`vegan`](https://github.com/vegandevs/vegan)         | Only dendrogram-based FD (`treedive()`)                     |`r pkg_has_tests("vegan")`    | :white_check_mark: | ![](https://www.r-pkg.org/badges/last-release/vegan)|              
| [`TPD`](https://github.com/cran/TPD)                  | FRic, FDiv, FEve but for probability distributions          |`r pkg_has_tests("TPD")`      | :x:                | ![](https://www.r-pkg.org/badges/last-release/TPD)|
