---
title: "3. Performance Benchmark with other packages"
output: rmarkdown::html_vignette
bibliography: performance.bib
vignette: >
  %\VignetteIndexEntry{3. Performance Benchmark with other packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- 
# Pre-render with 
knitr::knit("vignettes/_fundiversity_2-performance.Rmd", output = "vignettes/fundiversity_2-performance.Rmd")
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  class.warning = "collapsable-warnings",
  warning = FALSE,
  fig.path = "perf-",
  fig.height = 8.5,
  fig.width = 8,
  message = FALSE,
  error   = FALSE
)

suppressPackageStartupMessages({
  library("dplyr")
  library("ggplot2")
})
```

This vignette presents some performance tests ran between `fundiversity` and other functional diversity packages. Note that to avoid the dependency on other packages, this vignette is [**pre-computed**](https://ropensci.org/technotes/2019/12/08/precompute-vignettes/).

# Other packages to compute FD indices

## Main functions

Here is a table that summarizes the comparable functions (and their arguments) for functions included in `fundiversity`. Note that the package name is indicated before `::` followed by the function name.

|    Index Type    | Index Name                                        | Source                        | `fundiversity` function | Equivalent Functions |
|:-----------------|:--------------------------------------------------|:------------------------------|:------------------------|----------------------|
| α-diversity      | Functional Dispersion (FDis)                      | @Laliberte_distancebased_2010 | `fd_fdis()`             | ~~`BAT::dispersion()`~~ (doesn't compute FDis)<br />`FD::fdisp()`<br />~~`hillR::hill_func()`~~ (actually rely on `FD` for FDis computations and standardize the traits without telling the user)<br />`mFD::alpha.fd.multidim(..., ind_vect = "fdis")` |
| α-diversity      | Functional Divergence (FDiv)                      | @Villeger_New_2008            | `fd_fdiv()`             | `mFD::alpha.fd.multidim(..., ind_vect = "fdiv")`                                                                                                                                                                                                        |
| α-diversity      | Functional Evenness (FEve)                        | @Villeger_New_2008            | `fd_feve()`             | `mFD::alpha.fd.multidim(..., ind_vect = "feve")`                                                                                                                                                                                                        |
| α-diversity      | Functional Richness (FRic)                        | @Villeger_New_2008            | `fd_fric()`             | `BAT::alpha()` (tree, not strictly equal)<br />`BAT::hull.alpha()` (hull)<br />`mFD::alpha.fd.multidim(..., ind_vect = "fric")`                                                                                                                         |
| α-diversity      | Rao's Quadratic Entropy (Q)                       | @Villeger_Decomposing_2013    | `fd_raoq()`             | `adiv::QE()`<br />`BAT::rao()`<br />`hillR::hill_func()` (standardize traits without warning)<br />`mFD::alpha.fd.hill(..., q = 2, tau = "max")` (returns a slightly modified version of Q according to @Ricotta_Diversity_2009)                        |
| β-diversity      | Functional Richness Intersection (FRic_intersect) | @Rao_Diversity_1982           | `fd_fric_intersect()`   | `betapart::functional.beta.pair()`<br />`hillR::hill_func_parti_pairwise()`                                                                                                                                                                             |

The other packages are thus: `adiv` [@Pavoine_adiv_2020], `BAT` [@Cardoso_BAT_2015], `betapart` [@Baselga_betapart_2012], `FD` [@Laliberte_FD_2014], `hillR` [@Li_hillR_2018], and `mFD` [@Magneville_mFD_2022]. For fairness of comparison, even if `FD::dbFD()` contains most indices we're not considering it as it computes all indices together for each call, and would necessarily be slower.


## Benchmark between packages

We will now benchmark the functions included in `fundiversity` with the corresponding function in other packages using the `microbenchmark::microbenchmark()` function. We will use the fairly small (~220 species, 8 sites, 4 traits) provided dataset in `fundiversity`.

```{r initial}
tictoc::tic()  # Time execution of vignette
library(fundiversity)
data("traits_birds", package = "fundiversity")
data("site_sp_birds", package = "fundiversity")

dist_traits_birds <- dist(traits_birds)
```


### Functional Dispersion (FDis)

```{r bench-fdis}
fdis_bench <- microbenchmark::microbenchmark(
  fundiversity = {
    fundiversity::fd_fdis(traits_birds, site_sp_birds)
  },
  FD = {
    FD::fdisp(dist_traits_birds, site_sp_birds)
  },
  mFD = {
    mFD::alpha.fd.multidim(
    traits_birds, site_sp_birds, ind_vect = "fdis",
    scaling = FALSE, verbose = FALSE, details_returned = FALSE
  )
  },
  times = 30
)

ggplot2::autoplot(fdis_bench) +
  labs(title = "Functional Dispersion (FDis)")
```


### Functional Divergence (FDiv)

```{r bench-fdiv}
fdiv_bench <- microbenchmark::microbenchmark(
  fundiversity = fd_fdiv(traits_birds, site_sp_birds),
  mFD = mFD::alpha.fd.multidim(
    traits_birds, site_sp_birds, ind_vect = "fdiv",
    scaling = FALSE, verbose = FALSE
  ),
  times = 30
)

ggplot2::autoplot(fdiv_bench) +
  labs(title = "Functional Divergence (FDiv)")

```


### Functional Evenness (FEve)

```{r bench-feve}
feve_bench <- microbenchmark::microbenchmark(
  fundiversity = fd_feve(traits_birds, site_sp_birds),
  mFD = mFD::alpha.fd.multidim(
    traits_birds, site_sp_birds, ind_vect = "feve",
    scaling = FALSE, verbose = FALSE
  ),
  times = 30
)

ggplot2::autoplot(feve_bench) +
  labs(title = "Functional Evenness (FEve)")
```


### Functional Richness (FRic)

```{r bench-fric}
fric_bench <- microbenchmark::microbenchmark(
  fundiversity = fd_fric(traits_birds, site_sp_birds),
  BAT_tree = BAT::alpha(
    site_sp_birds, traits_birds
  ),
  BAT_alpha_hull = BAT::hull.alpha(
    BAT::hull.build(site_sp_birds, traits_birds)
  ),
  mFD = mFD::alpha.fd.multidim(
    traits_birds, site_sp_birds, ind_vect = "fric",
    scaling = FALSE, verbose = FALSE
  ),
  times = 30
)

ggplot2::autoplot(fric_bench) +
  labs(title = "Functional Richness (FRic)")
```


### Functional Richness Intersection (FRic_intersect)

```{r bench-fric-intersect}
fric_bench <- microbenchmark::microbenchmark(
  fundiversity  = fd_fric_intersect(traits_birds, site_sp_birds) ,
  betapart = betapart::functional.beta.pair(
    site_sp_birds, traits_birds
  ),
  hillR = hillR::hill_func_parti_pairwise(
    site_sp_birds, traits_birds, .progress = FALSE
  ),
  times = 30
)

ggplot2::autoplot(fric_bench) +
  labs(title = "Functional Richness Intersection (FRic)")
```


### Rao's Quadratic Entropy (Q)

```{r bench-raoq}
raoq_bench <- fric_bench <- microbenchmark::microbenchmark(
  fundiversity = fd_raoq(traits_birds, site_sp_birds),
  adiv= adiv::QE(
    site_sp_birds, dist_traits_birds
  ),
  BAT_rao           = BAT::rao(
    site_sp_birds, distance = traits_birds
  ),
  hillR_hill_func   = hillR::hill_func(
    site_sp_birds, traits_birds, fdis = FALSE
  ),
  mFD_alpha_fd_hill = mFD::alpha.fd.hill(
    site_sp_birds, dist_traits_birds, q = 2,
    tau = "max"
  ),
  times = 30
)

ggplot2::autoplot(raoq_bench) +
  labs(title = "Rao's Quadatric Entropy (Q)")
```



# Benchmark within `fundiversity`

We now proceed to the performance evaluation of functions within `fundiversity` with datasets of increasing sizes.

## Increasing the number of species

```{r bench-species-data}
make_more_sp <- function(n) {
  traits <- do.call(rbind, replicate(n, traits_birds, simplify = FALSE))
  row.names(traits) <- paste0("sp", seq_len(nrow(traits)))

  site_sp <- do.call(cbind, replicate(n, site_sp_birds, simplify = FALSE))
  colnames(site_sp) <- paste0("sp", seq_len(ncol(site_sp)))

  list(tr = traits, si = site_sp)
}

null_sp_1000   <- make_more_sp(5)
null_sp_10000  <- make_more_sp(50)
null_sp_100000 <- make_more_sp(500)
```

### Functional Richness

```{r bench-species-fric, fig.cap = "Performance comparison of `fd_fric()` with increasing number of species."}
bench_sp_fric <- microbenchmark::microbenchmark(
  species_200    = fd_fric(     traits_birds, site_sp_birds),
  species_1000   = fd_fric(  null_sp_1000$tr, null_sp_1000$si),
  species_10000  = fd_fric( null_sp_10000$tr, null_sp_10000$si),
  species_100000 = fd_fric(null_sp_100000$tr, null_sp_100000$si),
  times = 30
)

ggplot2::autoplot(bench_sp_fric)
bench_sp_fric
```

### Functional Divergence

```{r bench-species-fdiv, fig.cap = "Performance comparison of `fd_fdiv()` with increasing number of species."}
bench_sp_fdiv <- microbenchmark::microbenchmark(
  species_200    = fd_fdiv(     traits_birds, site_sp_birds),
  species_1000   = fd_fdiv(  null_sp_1000$tr, null_sp_1000$si),
  species_10000  = fd_fdiv( null_sp_10000$tr, null_sp_10000$si),
  species_100000 = fd_fdiv(null_sp_100000$tr, null_sp_100000$si),
  times = 30
)

ggplot2::autoplot(bench_sp_fdiv)
bench_sp_fdiv
```

### Rao's Quadratic Entropy

```{r bench-species-raoq, fig.cap = "Performance comparison of `fd_raoq()` with increasing number of species."}
bench_sp_raoq <- microbenchmark::microbenchmark(
  species_200    = fd_raoq(     traits_birds, site_sp_birds),
  species_1000   = fd_raoq(  null_sp_1000$tr, null_sp_1000$si),
  species_10000  = fd_raoq( null_sp_10000$tr, null_sp_10000$si),
  times = 30
)

ggplot2::autoplot(bench_sp_raoq)
bench_sp_raoq
```

### Functional Evenness

```{r bench-species-feve, fig.cap = "Performance comparison of `fd_feve()` with increasing number of species."}
bench_sp_feve <- microbenchmark::microbenchmark(
  species_200    = fd_feve(     traits_birds, site_sp_birds),
  species_1000   = fd_feve(  null_sp_1000$tr, null_sp_1000$si),
  species_10000  = fd_feve( null_sp_10000$tr, null_sp_10000$si),
  times = 15
)

ggplot2::autoplot(bench_sp_feve)
bench_sp_feve
```

### Comparing between indices

```{r bench-species-compare, fig.cap="Performance comparison between functions for distinct indices in `fundiversity` with increasing number of species. Smoothed trend lines and standard error envelopes ares shown."}
all_bench_sp <- list(fric = bench_sp_fric,
                     fdiv = bench_sp_fdiv,
                     raoq = bench_sp_raoq,
                     feve = bench_sp_feve) %>%
  bind_rows(.id = "fd_index") %>%
  mutate(n_sp = gsub("species_", "", expr) %>%
           as.numeric())

all_bench_sp %>%
  ggplot(aes(n_sp, time * 1e-9, color = fd_index)) +
  geom_point(alpha = 1/3) +
  geom_smooth() +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_brewer(type = "qual",
                     labels = c(fric = "FRic", fdiv = "FDiv", raoq = "Rao's Q",
                                feve = "FEve")) +
  labs(title = "Performance comparison between indices",
       x = "# of species", y = "Time (in seconds)",
       color = "FD index") +
  theme_bw() +
  theme(aspect.ratio = 1)
```

## Increasing the number of sites

```{r bench-sites-data}
make_more_sites <- function(n) {
  site_sp <- do.call(rbind, replicate(n, site_sp_birds, simplify = FALSE))
  rownames(site_sp) <- paste0("s", seq_len(nrow(site_sp)))

  site_sp
}

site_sp_100   <- make_more_sites(12)
site_sp_1000  <- make_more_sites(120)
site_sp_10000 <- make_more_sites(1200)
```


### Functional Richness

```{r bench-sites-fric, fig.cap = "Performance comparison of `fd_fric()` with increasing number of sites."}
bench_sites_fric <- microbenchmark::microbenchmark(
  sites_10    = fd_fric(traits_birds, site_sp_birds),
  sites_100   = fd_fric(traits_birds, site_sp_100),
  sites_1000  = fd_fric(traits_birds, site_sp_1000),
  sites_10000 = fd_fric(traits_birds, site_sp_10000),
  times = 15
)

ggplot2::autoplot(bench_sites_fric)
bench_sites_fric
```

### Functional Divergence

```{r bench-sites-fdiv, fig.cap = "Performance comparison of `fd_fdiv()` with increasing number of sites."}
bench_sites_fdiv <- microbenchmark::microbenchmark(
  sites_10    = fd_fdiv(traits_birds, site_sp_birds),
  sites_100   = fd_fdiv(traits_birds, site_sp_100),
  sites_1000  = fd_fdiv(traits_birds, site_sp_1000),
  sites_10000 = fd_fdiv(traits_birds, site_sp_10000),
  times = 15
)

ggplot2::autoplot(bench_sites_fdiv)
bench_sites_fdiv
```

### Rao's Quadratic Entropy

```{r bench-sites-raoq, fig.cap = "Performance comparison of `fd_raoq()` with increasing number of sites."}
bench_sites_raoq = microbenchmark::microbenchmark(
  sites_10    = fd_raoq(traits = NULL, site_sp_birds, dist_traits_birds),
  sites_100   = fd_raoq(traits = NULL, site_sp_100,   dist_traits_birds),
  sites_1000  = fd_raoq(traits = NULL, site_sp_1000,  dist_traits_birds),
  sites_10000 = fd_raoq(traits = NULL, site_sp_10000, dist_traits_birds),
  times = 15
)

ggplot2::autoplot(bench_sites_raoq)
bench_sites_raoq
```

### Functional Evenness

```{r bench-sites-feve, fig.cap = "Performance comparison of `fd_feve()` with increasing number of sites"}
bench_sites_feve <- microbenchmark::microbenchmark(
  sites_10    = fd_feve(traits = NULL, site_sp_birds, dist_traits_birds),
  sites_100   = fd_feve(traits = NULL, site_sp_100,   dist_traits_birds),
  sites_1000  = fd_feve(traits = NULL, site_sp_1000,  dist_traits_birds),
  sites_10000 = fd_feve(traits = NULL, site_sp_10000, dist_traits_birds),
  times = 15
)

ggplot2::autoplot(bench_sites_feve)
bench_sites_feve
```


### Comparing between indices

```{r bench-sites-compare, fig.cap = "Performance comparison between functions for distinct indices in `fundiversity` with increasing number of sites. Smoothed trend lines and standard error envelopes ares shown."}
all_bench_sites <- list(fric = bench_sites_fric,
                        fdiv = bench_sites_fdiv,
                        raoq = bench_sites_raoq,
                        feve = bench_sites_feve) %>%
  bind_rows(.id = "fd_index") %>%
  mutate(n_sites = as.numeric(gsub("sites", "", expr, fixed = TRUE)))

all_bench_sites %>%
  ggplot(aes(n_sites, time * 1e-9, color = fd_index)) +
  geom_point(alpha = 1/3) +
  geom_smooth() +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_brewer(type = "qual",
                     labels = c(fric = "FRic", fdiv = "FDiv", raoq = "Rao's Q",
                                feve = "FEve")) +
  labs(title = "Performance comparison between indices",
       x = "# of sites", y = "Time (in seconds)",
       color = "FD index") +
  theme_bw() +
  theme(aspect.ratio = 1)
```

<details>
<summary>Session info of the machine on which the benchmark was ran and time it took to run</summary>

```{r sessioninfo, echo = FALSE}
tictoc::toc(
  func.toc = function(x, y, msg) tictoc::toc.outmsg(
    x, y, paste0(msg, " seconds needed to generate this document")
    )
)
sessioninfo::session_info()
```

</details>

# References

```{js, echo=FALSE}
(function() {
  var codes = document.querySelectorAll('.collapsable-warnings');
  var code, i, d, s, p;
  for (i = 0; i < codes.length; i++) {
    code = codes[i];
    p = code.parentNode;
    d = document.createElement('details');
    s = document.createElement('summary');
    s.innerText = 'Warnings';
    // <details><summary>Details</summary></details>
    d.appendChild(s);
    // move the code into <details>
    p.replaceChild(d, code);
    d.appendChild(code);
  }
})();
```
