---
title: "2. Parallelize Computation of Indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Parallelize Computation of Indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- 
# Pre-render with 
knitr::knit("vignettes/_fundiversity_1-parallel.Rmd", output = "vignettes/fundiversity_1-parallel.Rmd")
-->

```{r vignette-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "vignettes/para-"
)
tictoc::tic()

options(fundiversity.memoise = FALSE)
```

Note: This vignette presents some performance tests ran between non-parallel and parallel versions of `fundiversity` functions. Note that to avoid the dependency on other packages, this vignette is [**pre-computed**](https://ropensci.org/technotes/2019/12/08/precompute-vignettes/).

Within `fundiversity` the computation of most indices can be parallelized using the `future` package. The goal of this vignette is to explain how to toggle and use parallelization in `fundiversity`. The functions that currently support parallelization are summarized in the table below:

```{r child="../man/rmdchunks/_fundiversity_functions.Rmd"}
```

Note that **memoization and parallelization cannot be used at the same time**. If the option `fundiversity.memoise` has been set to `TRUE` but the computation are parallelized, `fundiversity` will use unmemoised version of functions.

The `future` package provides a simple and general framework to allow asynchronous computation depending on the resources available for the user. The [first vignette of `future`](https://cran.r-project.org/package=future) gives a general overview of all its features. The main idea being that the user should write the code once and that it would run seamlessly sequentially, or in parallel on a single computer, or on a cluster, or distributed over several computers. `fundiversity` can thus run on all these different backends following the user's choice.

```{r setup}
library("fundiversity")

data("traits_birds", package = "fundiversity")
data("site_sp_birds", package = "fundiversity")
```

# Running code in parallel

By default the `fundiversity` code will run sequentially on a single core. To trigger parallelization the user needs to define a `future::plan()` object with a parallel backend such as `future::multisession` to split the execution across multiple R sessions.

```{r future-plan}
# Sequential execution
fric1 <- fd_fric(traits_birds)

# Parallel execution
future::plan(future::multisession)  # Plan definition
fric2 <- fd_fric(traits_birds)  # The code resolve in similar fashion

identical(fric1, fric2)
```

Within the `future::multisession` backend you can specify the number of cores on which the function should be parallelized over through the argument `workers`, you can change it in the `future::plan()` call:

```{r future-plan-workers, options}
future::plan(future::multisession, workers = 2)  # Only 2 cores are used
fric3 <- fd_fric(traits_birds)

identical(fric3, fric2)
```

To learn more about the different backends available and the related arguments needed, please refer to the documentation of `future::plan()` and the [overview vignette of `future`](https://cran.r-project.org/package=future).


# Performance comparison

We can now compare the difference in performance to see the performance gain thanks to parallelization:

```{r perf-comp}
future::plan(future::sequential)
non_parallel_bench <- microbenchmark::microbenchmark(
  non_parallel = {
    fd_fric(traits_birds)
  },
  times = 20
)

future::plan(future::multisession)
parallel_bench <- microbenchmark::microbenchmark(
  parallel = {
    fd_fric(traits_birds)
  },
  times = 20
)

rbind(non_parallel_bench, parallel_bench)
```

The non parallelized code runs faster than the parallelized one! Indeed, the parallelization in `fundiversity` parallelize the computation across different sites. So parallelization should be used when you have many sites on which you want to compute similar indices.

```{r make-more-sp}
# Function to make a bigger site-sp dataset
make_more_sites <- function(n) {
  site_sp <- do.call(rbind, replicate(n, site_sp_birds, simplify = FALSE))
  rownames(site_sp) <- paste0("s", seq_len(nrow(site_sp)))

  site_sp
}
```

For example with a dataset 5000 times bigger:

```{r bench-5000-times}
bigger_site <- make_more_sites(5000)

microbenchmark::microbenchmark(
  seq = { 
    future::plan(future::sequential) 
    fd_fric(traits_birds, bigger_site) 
  },
  multisession = { 
    future::plan(future::multisession, workers = 4)
    fd_fric(traits_birds, bigger_site) 
  },
  multicore = { 
    future::plan(future::multicore, workers = 4) 
    fd_fric(traits_birds, bigger_site) 
  }, times = 20
)
```

<details>
<summary>Session info of the machine on which the benchmark was ran and time it took to run</summary>

```{r sessioninfo, echo = FALSE}
tictoc::toc(
  func.toc = function(x, y, msg)
    tictoc::toc.outmsg(
      x, y, paste0(msg, " seconds needed to generate this document")
    )
)
sessioninfo::session_info()
```

</details>
