---
title: "Performance comparison with other functional diversity packages"
output: rmarkdown::html_vignette
bibliography: performance.bib
vignette: >
  %\VignetteIndexEntry{Performance comparison with other functional diversity packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  class.warning = "collapsable-warnings",
  warning = FALSE,
  fig.path = "fd-",
  fig.height = 8.5,
  fig.width = 8,
  message = FALSE
)

suppressPackageStartupMessages({
  library("dplyr")
  library("ggplot2")
})
```

This vignette presents some performance tests ran between `fundiversity` and other functional diversity packages. Note that to avoid the dependency on other packages, this vignette is [**pre-computed**](https://ropensci.org/technotes/2019/12/08/precompute-vignettes/).

# Other packages to compute FD indices

## Main functions

**`FD::dbFD()`** is probably the most known function to compute functional diversity indices. It comes from the `FD` package proposed by @Laliberte_FD_2014. It can compute many indices: functional richness, Rao's quadratic entropy, community-weighted mean, functional divergence, functional dispersion, and functional group richness. This swiss-army knife of functional diversity indices has many options and compute several indices in a single run.

**`hillR::hill_func()`** comes from the `hillR` package that computes Hill numbers for taxonomic, functional, and functional diversity [@Li_hillR_2018]. It implements indices proposed by @Chao_Unifying_2014. It computes Rao's quadratic entropy, the mean functional diversity per species, as well as the total functional diversity.

**`adiv::QE()`** comes from the `adiv` package which proposes a toolkit to analyze biodiversity [@Pavoine_adiv_2020]. This function computes Rao's quadratic entropy.

**`SYNCSA::rao.diversity()`** comes from `SYNCSA` package which proposes a simulation framework for meta-communities [@Debastiani_SYNCSA_2012]. The function computes Rao's quadratic entropy as well as functional redundancy.

## Benchmark between packages

We will now benchmark the functions included in `fundiversity` with the corresponding function in other packages using the `microbenchmark::microbenchmark()` function.

```{r setup}
tictoc::tic()
library(fundiversity)
data("traits_birds", package = "fundiversity")
data("site_sp_birds", package = "fundiversity")

dist_traits_birds <- dist(traits_birds)
```

With the fairly small (~220 species, 8 sites, 4 traits) provided dataset in `fundiversity`:

```{r bench-data-rao, fig.cap="Performance comparison in computing Rao's quadratic entropy between `fundiversity` and other packages"}
pkg_bench <- microbenchmark::microbenchmark(
  fundiversity = {
    fundiversity::fd_raoq(traits = NULL, site_sp_birds, dist_traits_birds)
  },
  adiv = {
    adiv::QE(site_sp_birds, dis = dist_traits_birds)
  },
  hillR = {
    hillR::hill_func(site_sp_birds, dist_traits_birds, traits_as_is = TRUE)[1, ]
  },
  SYNCSA = {
    SYNCSA::rao.diversity(site_sp_birds,
                          phylodist = as.matrix(dist_traits_birds),
                          standardize = FALSE)$PhyRao
  },
  FD = {
    suppressMessages(
      FD::dbFD(x = dist_traits_birds, a = site_sp_birds, stand.x = FALSE,
               calc.FRic = TRUE, calc.CWM = FALSE, calc.FDiv = FALSE,
               calc.FGR = FALSE, m = "max", stand.FRic = FALSE,
               messages = FALSE))
  },
  times = 30
)

ggplot2::autoplot(pkg_bench)
pkg_bench
```


# Benchmark within `fundiversity`

We now proceed to the performance evaluation of functions within `fundiversity` with datasets of increasing sizes.

## Increasing the number of species

```{r bench-species-data}
make_more_sp <- function(n) {
  traits <- do.call(rbind, replicate(n, traits_birds, simplify = FALSE))
  row.names(traits) <- paste0("sp", seq_len(nrow(traits)))

  site_sp <- do.call(cbind, replicate(n, site_sp_birds, simplify = FALSE))
  colnames(site_sp) <- paste0("sp", seq_len(ncol(site_sp)))

  list(tr = traits, si = site_sp)
}

null_sp_1000   <- make_more_sp(5)
null_sp_10000  <- make_more_sp(50)
null_sp_100000 <- make_more_sp(500)
```

### Functional Richness

```{r bench-species-fric, fig.cap = "Performance comparison of `fd_fric()` with increasing number of species."}
bench_sp_fric <- microbenchmark::microbenchmark(
  species_200    = fd_fric(     traits_birds, site_sp_birds),
  species_1000   = fd_fric(  null_sp_1000$tr, null_sp_1000$si),
  species_10000  = fd_fric( null_sp_10000$tr, null_sp_10000$si),
  species_100000 = fd_fric(null_sp_100000$tr, null_sp_100000$si),
  times = 30
)

ggplot2::autoplot(bench_sp_fric)
bench_sp_fric
```

### Functional Divergence

```{r bench-species-fdiv, fig.cap = "Performance comparison of `fd_fdiv()` with increasing number of species."}
bench_sp_fdiv <- microbenchmark::microbenchmark(
  species_200    = fd_fdiv(     traits_birds, site_sp_birds),
  species_1000   = fd_fdiv(  null_sp_1000$tr, null_sp_1000$si),
  species_10000  = fd_fdiv( null_sp_10000$tr, null_sp_10000$si),
  species_100000 = fd_fdiv(null_sp_100000$tr, null_sp_100000$si),
  times = 30
)

ggplot2::autoplot(bench_sp_fdiv)
bench_sp_fdiv
```

### Rao's Quadratric Entropy

```{r bench-species-raoq, fig.cap = "Performance comparison of `fd_raoq()` with increasing number of species."}
bench_sp_raoq <- microbenchmark::microbenchmark(
  species_200    = fd_raoq(     traits_birds, site_sp_birds),
  species_1000   = fd_raoq(  null_sp_1000$tr, null_sp_1000$si),
  species_10000  = fd_raoq( null_sp_10000$tr, null_sp_10000$si),
  times = 30
)

ggplot2::autoplot(bench_sp_raoq)
bench_sp_raoq
```

### Functional Evenness

```{r bench-species-feve, fig.cap = "Performance comparison of `fd_feve()` with increasing number of species."}
bench_sp_feve <- microbenchmark::microbenchmark(
  species_200    = fd_feve(     traits_birds, site_sp_birds),
  species_1000   = fd_feve(  null_sp_1000$tr, null_sp_1000$si),
  species_10000  = fd_feve( null_sp_10000$tr, null_sp_10000$si),
  times = 30
)

ggplot2::autoplot(bench_sp_feve)
bench_sp_feve
```

### Comparing between indices

```{r bench-species-compare, fig.cap="Performance comparison between functions for distinct indices in `fundiversity` with increasing number of species. Smoothed trend lines and standard error enveloppes ares shown."}
all_bench_sp <- list(fric = bench_sp_fric,
                     fdiv = bench_sp_fdiv,
                     raoq = bench_sp_raoq,
                     feve = bench_sp_feve) %>%
  bind_rows(.id = "fd_index") %>%
  mutate(n_sp = gsub("species_", "", expr) %>%
           as.numeric())

all_bench_sp %>%
  ggplot(aes(n_sp, time * 1e-9, color = fd_index)) +
  geom_point(alpha = 1/3) +
  geom_smooth() +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_brewer(type = "qual",
                     labels = c(fric = "FRic", fdiv = "FDiv", raoq = "Rao's Q",
                                feve = "FEve")) +
  labs(title = "Performance comparison between indices",
       x = "# of species", y = "Time (in seconds)",
       color = "FD index") +
  theme_bw() +
  theme(aspect.ratio = 1)
```

## Increasing the number of sites

```{r bench-sites-data}
make_more_sites <- function(n) {
  site_sp <- do.call(rbind, replicate(n, site_sp_birds, simplify = FALSE))
  rownames(site_sp) <- paste0("s", seq_len(nrow(site_sp)))

  site_sp
}

site_sp_100   <- make_more_sites(12)
site_sp_1000  <- make_more_sites(120)
site_sp_10000 <- make_more_sites(1200)
```


### Functional Richness

```{r bench-sites-fric, fig.cap = "Performance comparison of `fd_fric()` with increasing number of sites."}
bench_sites_fric = microbenchmark::microbenchmark(
  sites_10    = fd_fric(traits_birds, site_sp_birds),
  sites_100   = fd_fric(traits_birds, site_sp_100),
  sites_1000  = fd_fric(traits_birds, site_sp_1000),
  sites_10000 = fd_fric(traits_birds, site_sp_10000),
  times = 30
)

ggplot2::autoplot(bench_sites_fric)
bench_sites_fric
```

### Functional Divergence

```{r bench-sites-fdiv, fig.cap = "Performance comparison of `fd_fdiv()` with increasing number of sites."}
bench_sites_fdiv = microbenchmark::microbenchmark(
  sites_10    = fd_fdiv(traits_birds, site_sp_birds),
  sites_100   = fd_fdiv(traits_birds, site_sp_100),
  sites_1000  = fd_fdiv(traits_birds, site_sp_1000),
  sites_10000 = fd_fdiv(traits_birds, site_sp_10000),
  times = 30
)

ggplot2::autoplot(bench_sites_fdiv)
bench_sites_fdiv
```

### Rao's Quadratic Entropy

```{r bench-sites-raoq, fig.cap = "Performance comparison of `fd_raoq()` with increasing number of sites."}
bench_sites_raoq = microbenchmark::microbenchmark(
  sites_10    = fd_raoq(traits = NULL, site_sp_birds, dist_traits_birds),
  sites_100   = fd_raoq(traits = NULL, site_sp_100,   dist_traits_birds),
  sites_1000  = fd_raoq(traits = NULL, site_sp_1000,  dist_traits_birds),
  sites_10000 = fd_raoq(traits = NULL, site_sp_10000, dist_traits_birds),
  times = 30
)

ggplot2::autoplot(bench_sites_raoq)
bench_sites_raoq
```

### Functional Evenness

```{r bench-sites-feve, fig.cap = "Performance comparison of `fd_feve()` with increasing number of sites"}
bench_sites_feve = microbenchmark::microbenchmark(
  sites_10    = fd_feve(traits = NULL, site_sp_birds, dist_traits_birds),
  sites_100   = fd_feve(traits = NULL, site_sp_100,   dist_traits_birds),
  sites_1000  = fd_feve(traits = NULL, site_sp_1000,  dist_traits_birds),
  sites_10000 = fd_feve(traits = NULL, site_sp_10000, dist_traits_birds),
  times = 30
)

ggplot2::autoplot(bench_sites_feve)
bench_sites_feve
```


### Comparing between indices

```{r bench-sites-compare, fig.cap = "Performance comparison between functions for distinct indices in `fundiversity` with increasing number of sites. Smoothed trend lines and standard error enveloppes ares shown."}
all_bench_sites <- list(fric = bench_sites_fric,
                        fdiv = bench_sites_fdiv,
                        raoq = bench_sites_raoq,
                        feve = bench_sites_feve) %>%
  bind_rows(.id = "fd_index") %>%
  mutate(n_sites = gsub("sites", "", expr) %>%
           as.numeric())

all_bench_sites %>%
  ggplot(aes(n_sites, time * 1e-9, color = fd_index)) +
  geom_point(alpha = 1/3) +
  geom_smooth() +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_brewer(type = "qual",
                     labels = c(fric = "FRic", fdiv = "FDiv", raoq = "Rao's Q",
                                feve = "FEve")) +
  labs(title = "Performance comparison between indices",
       x = "# of sites", y = "Time (in seconds)",
       color = "FD index") +
  theme_bw() +
  theme(aspect.ratio = 1)
```

<details>
<summary>Session info of the machine on which the benchmark was ran and time it took to run</summary>

```{r sessioninfo, echo = FALSE}
tictoc::toc(func.toc = function(x, y, msg) tictoc::toc.outmsg(x, y, paste0(msg, " seconds needed to generate this document")))
sessioninfo::session_info()
```

</details>

# References

```{js, echo=FALSE}
(function() {
  var codes = document.querySelectorAll('.collapsable-warnings');
  var code, i, d, s, p;
  for (i = 0; i < codes.length; i++) {
    code = codes[i];
    p = code.parentNode;
    d = document.createElement('details');
    s = document.createElement('summary');
    s.innerText = 'Warnings';
    // <details><summary>Details</summary></details>
    d.appendChild(s);
    // move the code into <details>
    p.replaceChild(d, code);
    d.appendChild(code);
  }
})();
```
