---
title: "Verification that indices are implemented correctly"
output: rmarkdown::html_vignette
link-citations: yes
bibliography: fundiversity.bib
vignette: >
  %\VignetteIndexEntry{Verification that indices are implemented correctly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r vignette-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Most research results nowadays rely heavily on results generated by
computational software. Yet, this software, while occupying an increasingly
central place in science, is often poorly tested. Users are never presented any
arguments about why they should trust the code provided to them.

We want to contribute to change this trend and in this vignette, we show that 
indices computed by `fundiversity` behave as theoretically expected, and as
presented in @Villeger_New_2008.

```{r load-pkg}
library(fundiversity)
```

For this, we reproduce the Figure 2 (including all of their facets) of their 
article. If you re-use this figure, please cite the original article.

We start by reproducing Figure 2 a, to check that the values computed by
`fundiversity` are correct. These are the data found in Figure 2 of the article:

```{r panel-a}
data_a <- matrix(byrow = TRUE, ncol = 2,
  c(
    0.0, 1.0,
    0.5, 1.0,
    1.0, 1.0,
    1.5, 1.0,
    2.0, 1.0,
    1.0, 0.0,
    1.0, 0.5,
    1.0, 1.5,
    1.0, 2.0
  )
)
rownames(data_a) <- paste0("species", seq_len(nrow(data_a)))
print(data_a)
```

```{r plot-data-a, out.width = '100%', fig.width=7, fig.asp=1}
plot(data_a, pch = 19, asp = 1, xlab = "Trait 1", ylab = "Trait 2")
```

We then compute the functional diversity indices.

```{r compute-a}
fric_a <- fd_fric(data_a)[["FRic"]]
feve_a <- fd_feve(data_a)[["FEve"]]
fdiv_a <- fd_fdiv(data_a)[["FDiv"]]
fdis_a <- fd_fdis(data_a)[["FDis"]]
raoq_a <- fd_raoq(data_a)[["Q"]]
```

As expected, we found an FRic value of `r fric_a` (should be 2), an FEve value
of `r feve_a` (should be 1), and a FDiv value of `r round(fdiv_a, 3)`
(should be 0.692).

# Changes of abundance

If we change the abundance, but not the coordinates (i.e., not the species
characteristics), we expect changes in FEve and FDiv but not FRic. 

More precisely, equal increases in abundance for the species at the vertices of the convex hull don't influence FEve, but lead to a higher FDiv. These correspond to
the data found in panel c of the Figure 2.

```{r data-c}
l <- 2 # common species
s <- 1 # rare species

wc <- matrix(c(l, s, l, s, l, l, s, s, l), nrow = 1)
colnames(wc) <- rownames(data_a)
rownames(wc) <- "site"
```

We compute again the functional diversity indices:

```{r compute-c}
fric_c <- fd_fric(data_a)[["FRic"]]
feve_c <- fd_feve(data_a, wc)[["FEve"]]
fdiv_c <- fd_fdiv(data_a, wc)[["FDiv"]]
fdis_c <- fd_fdis(data_a, wc)[["FDis"]]
raoq_c <- fd_raoq(data_a, wc)[["Q"]]
```

As expected, we found an FRic value of `r fric_c` (should be 2), an FEve value
of `r feve_c` (should be 1), and a FDiv value of `r round(fdiv_c, 3)`
(should be 0.714).

Conversely, changes of abundances on a single trait axis don't impact FDiv but
reduce FEve (this correspond to the panel b of Figure 2):

```{r data-b}
l <- 2 # common species
s <- 1 # rare species

wb <- matrix(c(l, l, l, l, l, s, s, s, s), nrow = 1)
colnames(wb) <- rownames(data_a)
rownames(wb) <- "site"
```

We can compute weighted functional diversity indices:

```{r compute-b}
fric_b <- fd_fric(data_a)[["FRic"]]
feve_b <- fd_feve(data_a, wb)[["FEve"]]
fdiv_b <- fd_fdiv(data_a, wb)[["FDiv"]]
fdis_b <- fd_fdis(data_a, wb)[["FDis"]]
raoq_b <- fd_raoq(data_a, wb)[["Q"]]
```

As expected, we found an FRic value of `r fric_b` (should be 2), an FEve value
of `r feve_b` (should be 0.778), and a FDiv value of `r round(fdiv_b, 3)`
(should be 0.692).

# Changes of coordinates (= traits)

We now change the species characteristics (i.e., the coordinates) instead of
the abundances. The changes of coordinates without changing abundances can
affect all diversity indices. If the coordinates are only changed for species
that are not on the convex hull would only affect FEve and FDiv.

For example, we can change the coordinates of species to be at the same distance
of the centroid of the species, so that it doesn't affect the value of FDiv, but
put them further away from points on the convex hull. This would decrease FEve
because species would be spaced less evenly in the space (it corresponds to
panel d of Figure 2):

```{r data-d}
shift <- 1/(2*sqrt(2))

data_d <- matrix(c(
  1-shift, 1-shift,
  1-shift, 1+shift,
  1+shift, 1-shift,
  1+shift, 1+shift,
  1.00 , 0.00 ,
  2.00 , 1.00 ,
  1.00 , 2.00 ,
  0.00 , 1.00 ,
  1.00 , 1.00),
  byrow = TRUE,
  ncol = 2
)

```

```{r compute-d}
fric_d <- fd_fric(data_d)[["FRic"]]
feve_d <- fd_feve(data_d)[["FEve"]]
fdiv_d <- fd_fdiv(data_d)[["FDiv"]]
fdis_d <- fd_fdis(data_d)[["FDis"]]
raoq_d <- fd_raoq(data_d)[["Q"]]
```

As expected, we found an FRic value of `r fric_d` (should be 2), an FEve value
of `r round(feve_d, 3)` (should be 0.891), and a FDiv value of `r round(fdiv_d, 3)`
(should be 0.692).

We can also change the coordinates of species (their traits) to only affect FDiv and not FEve by, instead, moving species on the convex hull while keeping their distance
to other points equal. This corresponds to the panel e Figure 2 of the paper:

```{r data-e}
data_e <- matrix(c(
  0.0, 1.0,
  0.5, 0.5,
  1.0, 1.0,
  1.5, 0.5,
  2.0, 1.0,
  1.0, 0.0,
  0.5, 1.5,
  1.5, 1.5,
  1.0, 2.0),
  byrow = TRUE,
  ncol = 2
)
```

We can then compute functional diversity indices:

```{r compute-e}
fric_e <- fd_fric(data_e)[["FRic"]]
feve_e <- fd_feve(data_e)[["FEve"]]
fdiv_e <- fd_fdiv(data_e)[["FDiv"]]
fdis_e <- fd_fdis(data_e)[["FDis"]]
raoq_e <- fd_raoq(data_e)[["Q"]]
```

As expected, we found an FRic value of `r fric_e` (should be 2), an FEve value
of `r feve_e` (should be 1), and a FDiv value of `r round(fdiv_e, 2)`
(should be 0.78).
