---
title: "Performance comparison with other functional diversity packages"
output: rmarkdown::html_vignette
bibliography: performance.bib
vignette: >
  %\VignetteIndexEntry{Performance comparison with other functional diversity packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette presents some performance tests ran between `fundiversity` and other functional diversity packages. Note that to avoid the dependency on other packages, this vignette is **pre-computed**.

# Other functions/packages to compute functional diversity indices

**`FD::dbFD()`** is probably the most known function to compute functional diversity indices. It comes from the `FD` package proposed by Laliberté and Legendre [-@Laliberte_FD_2014]. It can compute many indices: functional richness, Rao's quadratic entropy, community-weighted mean, functional divergence, functional dispersion, and functional group richness. This swiss-army knife of functional diversity indices has many options and compute several indices in a single run.

**`hillR::hill_func()`** comes from the `hillR` package that computes Hill numbers for taxonomic, functional, and functional diversity [@Li_hillR_2018]. It implements indices proposed by Chao and colleagues [-@Chao_Unifying_2014]. It computes Rao's quadratic entropy, the mean functional diversity per species, the toal functional diversity.

**`adiv::QE()`** comes from the `adiv` package which proposes a toolkit to analyze biodiversity [@Pavoine_adiv_2020]. This function computes Rao's quadratic entropy.

# Test data

We will now benchmark the functions included in `fundiversity` with the corresponding function in other packages.

```{r setup}
library(fundiversity)
```

```{r bench-traits}

```

```{r bench-species}

```

```{r bench-both}

```

<details>
<summary>Session Info on which the benchmark was ran</summary>
```{r sessioninfo}
sessioninfo::session_info()
```
</details>

# References
