---
title: "Performance comparison with other functional diversity packages"
output: rmarkdown::html_vignette
bibliography: performance.bib
vignette: >
  %\VignetteIndexEntry{Performance comparison with other functional diversity packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette presents some performance tests ran between `fundiversity` and other functional diversity packages. Note that to avoid the dependency on other packages, this vignette is **pre-computed**.

# Other functions/packages to compute functional diversity indices

**`FD::dbFD()`** is probably the most known function to compute functional diversity indices. It comes from the `FD` package proposed by Laliberté and Legendre [-@Laliberte_FD_2014]. It can compute many indices: functional richness, Rao's quadratic entropy, community-weighted mean, functional divergence, functional dispersion, and functional group richness. This swiss-army knife of functional diversity indices has many options and compute several indices in a single run.

**`hillR::hill_func()`** comes from the `hillR` package that computes Hill numbers for taxonomic, functional, and functional diversity [@Li_hillR_2018]. It implements indices proposed by Chao and colleagues [-@Chao_Unifying_2014]. It computes Rao's quadratic entropy, the mean functional diversity per species, as well as the total functional diversity.

**`adiv::QE()`** comes from the `adiv` package which proposes a toolkit to analyze biodiversity [@Pavoine_adiv_2020]. This function computes Rao's quadratic entropy.

**`SYNCSA::rao.diversity()`** comes from `SYNCSA` package which proposes a simulation framework for meta-communities [@Debastiani_SYNCSA_2012]. The function computes Rao's quadratic entropy as well as functional redundancy.

# Benchmark

We will now benchmark the functions included in `fundiversity` with the corresponding function in other packages using the `microbenchmark::microbenchmark()` function.

```{r setup}
library(fundiversity)
data("traits_birds", package = "fundiversity")
data("site_sp_birds", package = "fundiversity")

dist_traits_birds <- dist(traits_birds)
```

With the fairly small (~220 species, 8 sites, 4 traits) provided dataset in `fundiversity`:

```{r bench-data-rao}
rao_bench <- microbenchmark::microbenchmark(
  fundiversity = {
    fundiversity::fd_raoq(traits = NULL, site_sp_birds, dist_traits_birds)
  },
  adiv = {
    adiv::QE(site_sp_birds, dis = dist_traits_birds)
  },
  hillR = {
    hillR::hill_func(site_sp_birds, dist_traits_birds, traits_as_is = TRUE)[1, ]
  },
  SYNCSA = {
    SYNCSA::rao.diversity(site_sp_birds,
                          phylodist = as.matrix(dist_traits_birds),
                          standardize = FALSE)$PhyRao
  },
  times = 1000
)

ggplot2::autoplot(rao_bench)

knitr::kable(summary(rao_bench))
```


## Increasing the number of species

```{r bench-species}
mega_trait_birds
```

## Increasing the number of sites

```{r}

```

## Increasing both

```{r bench-both}

```

<details>
<summary>Session info of the machine on which the benchmark was ran</summary>
```{r sessioninfo}
sessioninfo::session_info()
```
</details>

# References
